---
title: "P8106 Data ScienceII Homework2"
author: "Yueran Zhang(yz4188)"
date: '2023-03-05'
output:
  pdf_document:
    toc_depth: 2
  html_document:
    toc_depth: '2'
---

In this exercise, we build nonlinear models using the “College” data. The dataset contains statistics for 565 US Colleges from a previous issue of US News and World Report. The response variable is the out-of-state tuition (Outstate).Partition the dataset into two parts: training data (80%) and test data (20%).

# R Package
```{r R package, echo = T, message = FALSE, results='hide', warning=FALSE}
library(caret)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
```

# Import Dataset

```{r Dataset Import,echo = T, message = FALSE, results='hide', warning=FALSE}

set.seed(123)

# Load dataset + clean data
College = read.csv("/Users/yueranzhang/Desktop/DSII/HW2/DataSet/College.csv")[-1] %>% 
janitor::clean_names() %>% 
na.omit()

# Data Partition
RowTrain <- createDataPartition(y = College$outstate,
                                p = 0.8,
                                list = FALSE)
train_data <- College[RowTrain,]
test_data <- College[-RowTrain,]
```

\newpage

(a) Fit smoothing spline models using perc.alumni as the only predictor of Outstate for a range of degrees of freedom, as well as the degree of freedom obtained by generalized cross-validation, and plot the resulting fits. Describe the results obtained.

```{r QuestionA}

set.seed(123)

perc_alumni.grid <- seq(from =0, to = 70, by = 1)
fit.ss <- smooth.spline(train_data$perc_alumni, train_data$outstate,cv = TRUE)

fit.ss$df

fit.ss$lambda

# plot the fit

pred.ss <- predict(fit.ss,
                   x =perc_alumni.grid)

pred.ss.df <- data.frame(pred = pred.ss$y,
                         perc_alumni = perc_alumni.grid)

p <- ggplot(data = train_data, aes(x = perc_alumni, y = outstate)) +
     geom_point(color = rgb(.2, .4, .2, .5))

p +
geom_line(aes(x = perc_alumni.grid, y = pred), data = pred.ss.df,
          color = rgb(.8, .1, .1, 1)) + theme_bw()
```

* The degree of freedom that we obtained by generalized cross-validation is `r fit.ss$df`. From the above plot, the smoothing spline is nearly to a linear line, or that's to say this model fits the data quite well.
* It is noticed that as we get the $\lambda$ value is `r fit.ss$lambda`, which is pretty large, so for the function estimate  $f_\lambda$ is essentially constrained to have a zero penalty, and it is forced to be smoother.



